Solving the Poisson-equation in one dimension: \\ \normalsize{TDMA vs LU} \\ \hrulefill\small{ FYS3150: Computational Physics }\hrulefill

### 2016-09-16 Project1/article/article.tex
Introduction and abstract

\title[Solving the Poisson-equation in one dimension]{Solving the Poisson-equation in one dimension \\
  \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}

@@

\begin{titlepage}
\begin{abstract}
Lorem ipsum dolor sit amet, consectur adipiscing lit. Nullam ut lacus eget lorem...
\end{abstract}
\maketitle
\tableofcontents
\end{titlepage}

\section{Introduction}

\section{Theory}
\subsection{The Poisson Equation}

The Poisson equation is a classical equation from electromagnetism. The electrostatic potential $\Phi$ is generated by a localized charge distribution $\rho(\vb{r})$. In three dimensions the equation reads
\begin{equation}
\laplacian\Phi=-4\pi \rho(\vb{r})
\end{equation}
where $\laplacian$ is the Laplace operator. In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study I am assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem. only dependent on radius $r$.
\begin{equation}
\laplacian=\frac{1}{r^2}\frac{d}{dr}\left(r^2\frac{d\Phi}{dr} \right)
\end{equation}
By substituting $\Phi(r)=\phi(r)/r$ the Poisson equation is reduced to 
\begin{equation}
\frac{d^2\phi}{dr^2}=-4\pi r\rho(r)
\end{equation}
and by letting $\phi \rightarrow u$ and $r \rightarrow x$ one is left with the very simple equation
\begin{equation}
-u''(x)=f(x) \label{eq:2nd}
\end{equation}

@@

\section{Algorithms}
Two main methods are implemented and compared. The first method is
gaussian elimination of the tridiagonal matrix $A$, also known as the
\emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
elimination that can ve used to solve tridiagonal systems of
equations. The method is improved upon in order to take into account
the fact that the matrix we are dealing with has the same numbers
along the diagonals. The second method is the LU-decomposistion
method.

@@

\section{Discussion}
When choosing a suitible value for $n$, one must take into account both precision and speed. Our intuition tells us that a larger $n$ should result in a higher precision for the numerical derivative. We can conclude from figure \ref{fig:relerror} that this is only true up to a certain point, because of loss in numerical representation accuracy within the computer. An ideal n will be in the range of $n\approx 10^5$ which is equivalent to $h=10^-5$. Increasing this value will not increase run time dratically for the tridiagnoal matrix algorithm. Choosing $n=10^5$ leaves us with the ability to calculate with a high certainty of 5 decimals points and with low memory cost.

>>>>

\title[Solving the Poisson-equation in one dimension]{Solving the Poisson-equation in one dimension: \\
\normalsize{TDMA vs LU} \\
  \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}

@@

\begin{titlepage}
\begin{abstract}
A one-dimensional version of the Poisson equation with Dirichlet boundary conditions is solved using two different algorithms. The two algorithms employed are the tridiagonal matrix algorithm and the LU decomposition method. We find that a fine-tuned version of the tridiagonal matrix algorithm is $10^5$ times faster than the general LU decomposition method and $10$ times as precise.
\end{abstract}
\maketitle
\tableofcontents
\end{titlepage}

\section{Introduction}
In this study the one-dimensional Poisson dimension will be solved using two different algorithms. The Poisson equation is a partial differential equation fo elliptic type with broad application. In physics it is for instance used to describe the potential field of a charge, or mass density distribution. By assuming spherical symmetry, the Poisson equation is reduced to one dimension.

Herein, we devise two algorithms for solving the Poisson equation. The first one is the tridiagonal matrix algorithm, and the second is the LU decomposition method. We see how the tridiagonal matrix algorithm can be tailored to the problem and therefore becomes vastly more efficient than the more general LU method. The algorithms are tested at different precision levels.

Firstly, the physical background for the Poisson equation and how the second derivative can be approximated by a three point formula is presented. Secondly, we provide a thorough elaboration of how the algorithms can be implemented. Thirdly, results of the algorithms perfomance is presented. Lastly, we provide a discussion of the findings and some concluding remarks.

\section{Theory}
\subsection{The Poisson Equation}
The Poisson equation is a classical equation from electromagnetism. The electrostatic potential $\Phi$ is generated by a localized charge distribution $\rho(\vb{r})$. In three dimensions the equation reads
\begin{equation}
\laplacian\Phi=-4\pi \rho(\vb{r})
\end{equation}
where $\laplacian$ is the Laplace operator. In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study I am assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem. only dependent on radius $r$.
\begin{equation}
\laplacian=\frac{1}{r^2}\frac{d}{dr}\left(r^2\frac{d\Phi}{dr} \right)
\end{equation}
By substituting $\Phi(r)=\phi(r)/r$ the Poisson equation is reduced to 
\begin{equation}
\frac{d^2\phi}{dr^2}=-4\pi r\rho(r)
\end{equation}
and by letting $\phi \rightarrow u$ and $r \rightarrow x$ one is left with the very simple equation
\begin{equation}
-u''(x)=f(x) \label{eq:2nd}
\end{equation}

@@

\section{Algorithms}
Two main methods are implemented and compared. The first method is
gaussian elimination of the tridiagonal matrix $A$, also known as the
\emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
elimination that can ve used to solve tridiagonal systems of
equations. The method is improved upon in order to take into account
the fact that the matrix we are dealing with has the same numbers
along the diagonals. The second method is the LU-decomposistion
method. Code snippets are included in this section, for the full programs follow the link to github at the first page of this article.

@@

\section{Discussion}
When choosing a suitible value for $n$, one must take into account both precision and speed. Our intuition tells us that a larger $n$ should result in a higher precision for the numerical derivative. We can conclude from figure \ref{fig:relerror} that this is only true up to a certain point, because of loss in numerical representation accuracy within the computer. An ideal n will be in the magnitude of $10^5$ which is equivalent to $h=10^-5$. Increasing this value will not increase run time dratically for the tridiagnoal matrix algorithm. Choosing $n=10^5$ gives us the ability to calculate with a high certainty of 5 decimals points and with low memory cost.
